Ionization of molecules by multicharged bared ions by using the stoichiometric model

### 2019-09-06 NewMonti24.tex
added binding energy spectra and corrections

The first attempt to develop a comprehensive but straightforward 
phenomenological model for electron ejection from large molecules was 
proposed by Toburen {\it et al.}~\cite{toburen1975,toburen1976}. 
The authors found it convenient to scale the experimental ionization 
cross section in terms of the number of outer or weakly bound valence 
electrons (total number of electrons minus the inner-shell ones). 
We introduce the cross section per weakly bound electron or ionization 
cross section per valence electron, $\sigma_{e}$, as
\begin{equation}
\sigma_{e}=\frac{\sigma_{M}}{N_{M}}=\frac{\sum\limits_{\alpha}
n_{\alpha}\sigma_{\alpha}}{\sum\limits_{\alpha}n_{\alpha}\nu_{\alpha}}
=\sigma_{e}(v)\,, 
\label{27} 
\end{equation}
where
\begin{equation}
\nu_{\alpha}=\left\{ 
\begin{array}{ll}
4, & \text{for C,} \\ 
5, & \text{for N and P,} \\ 
6, & \text{for O and S,}
\end{array}\right.
\label{eq:nelec} 
\end{equation}
and $\nu_{\text{H}}=1$. The Toburen rule can be stated by saying that 
$\sigma_{e}$ is a \textit{universal} parameter independent on the 
molecule, which depends solely on the impact velocity. 
In Ref.~\cite{toburen1976}, it was found experimentally that for proton 
impact on some simple molecules, $\sigma_{e}$ results
\begin{equation}
\begin{array}{cccc}
E\text{(MeV)}               & 0.25 & 1.00 & 2.00 \\ 
\sigma_{e} (10^{-16}cm^{2}) & 0.39 & 0.17 & 0.11
\end{array}
\label{30}
\end{equation}
with an estimated error of about 20\%. A similar ratio was found in 
Ref.~\cite{itoh2013} for proton impact on uracil and adenine. 
Figure 4a shows $\sigma_{e}$ calculated with the CDW as a function of 
the impact velocity for different projectile charges computed with the 
SSM for the sixteen molecular targets displayed in 
Table~\ref{tab:families}. The universality with $Z$ is the one provided 
by Born approximation, i.e., $\sigma _{e}(Z)=Z^{2}\sigma_{e}(Z=1)$, 
and it holds for large impact velocities, as shown Figure 4a.
Of course, for lower impact velocities, the CDW breaks the behavior of 
the $Z^{2}$ rule. Although the Toburen rule holds for high energies, 
its performance is still not satisfactory: the universal band is quite 
broad. The departure of our theoretical 
results from the Toburen rule can be easily explained by the fact that 
our atomic cross sections $\sigma_{\alpha}$ behave differently for 
each atom in terms of the projectile charges. By inspecting Figure 1, 
one can easily see that the rule $\sigma_{\alpha}/v_{\alpha}\sim
\sigma_{e}$ is not well satisfied by the CDW. A much better general 
atomic rule is given by
\begin{equation}
\sigma_{e}^{\text{CDW}}=\frac{\sum\limits_{\alpha}
n_{\alpha}\sigma_{\alpha}^{\text{CDW}}}{\sum\limits_{\alpha}n_{\alpha}
\nu_{\alpha}^{\text{CDW}}},
\label{32} 
\end{equation}
where 
\begin{equation}
\nu_{\alpha }^{\text{CDW}} \sim\left\{ 
\begin{array}{ll}
4,\ \ \  & \text{for C, N, and O,} \\ 
4.5,\ \  & \text{for P and S}
\end{array}
\right. 
\label{35}
\end{equation}
and, obviously, $\nu_{\text{H}}^{\text{CDW}}=1$. The cross sections 
$\sigma_{e}^{\text{CDW}}$ are plotted in Figure~4b. A much better sharp 
band is observed, especially at high energy, where the theory is 
expected to work. It will be interesting for experiments to cross-check 
this theoretical prediction.

@@

In Figure 7, we display the relative difference of ionization cross 
sections 
\begin{equation}
e=\frac{\sigma^{\prime}-\sigma}{\sigma}\times 100
\label{120}
\end{equation}
for the four DNA basis. The difference obtained is of just very few 
percents, which indicates that the SSM is a quite robust model to treat 
the type of molecule we are dealing with, and of course within the 
range error that one can expect from the SSM.

We now return to the calculation of the mean electron energy of H$^{+}$ and
He$^{2+~}$ in water\ \ $\overline{E}_{H_{2}O}$. If we consider that \ the
residual charge of \ each hydrogen in water is $q_{H}=$+0.35 \ \cite%
{rappe1991}\ which implies that on Oxygen we have \ a residual charge of $%
q_{O}=$-0.7, we could re-write the formula of water using (\ref{120}) as H$%
_{1.3}$O$_{1.17}$ considering $n_{O}^{CDW}\simeq 4$ and $n_{H}^{CDW}\simeq
1. $ Recalculating mean electron energy by impact of \ H$^{+}$ and He$^{2+~}$
on water at 1 MeV/amu we obtain: $\overline{E}_{H_{2}O}=$46.2 \ and 47.6 eV
which compared a little better with the experiments: 51.5 and 52.2 eV \cite%
{pimblott2007}, respectively. i.e. within 10\% of the experiments which is
not bad accounting for the simplicity of our method.

Another parameter that we can calculate in our model is the fraction of
energy carried by the electron $f$ defined as%
\begin{equation}
f=\frac{\sum\limits_{\alpha }n_{\alpha }\sigma _{\alpha }\overline{%
E_{\alpha }}}{\sum\limits_{\alpha }n_{\alpha }(\sigma _{\alpha }\overline{%
E_{\alpha }}+\sum\limits_{nl}\sigma _{\alpha nl}E_{\alpha nl})}  \label{90}
\end{equation}%
Using our shell-to-shell ionization cross sections $\sigma _{\alpha nl}\ $%
\cite{miraglia2019}\ and atomic binding energies of the atom$\ \alpha ,$ $%
E_{\alpha nl}\ $\cite{clementi}, we obtain $f=$0$.70\ $and $\ 0.71$ for \ H$%
^{+}$ and He$^{2+~}$ on water at 1 MeV/amu,. which differ from the
experiments \ $f=0.81$ and $0.80$, \ respectively.

This difference with the experiment could be attributed to the difference of
binding energy between the molecular water 12.65 eV and the atomic component
H (13.6 eV) and O (17.19 eV), which is an issue beyond our independent atom
approximation and so our model cannot not accounted for. We could estimate
this contribution resorting, once more to the stoichiometric, model, we can
arriesgar a upper-limit difference per molecule of water\ as

\begin{equation}
\delta E=(13.6-12.65)1.17+(17.19-12.65)1.3=7\ \text{eV}  \label{80}
\end{equation}%
Most of this energy should be transfer to the kinetics energy of the emitted
electron. Thus we can \ add 7 eV to the previous values to give $\overline{E}%
_{H_{2}O}=$53.2 \ and 54.6 eV for \ H$^{+}$ and He$^{2+~}$ on water at 1
MeV/amu, respectively, much near to the experiments: 51.5 and 52.2 eV. If we
add $\delta E$ to the numerator of Eq.(\ref{90}) we obtain\ $f=0.81$ and $%
0.80$ in perfect agreement with the experiments (see Table 1 of Ref.(\cite%
{pimblott2007})

\section{CONCLUSIONS}

We have calculated ionization cross sections by impact of antiprotons, H$^{+}
$, He$^{2+}$, Be$^{4+}$, C$^{6+}$,\ and O$^{8+}$ with molecule involved in
biological basis containing H, C, N, O, P and S with the CDW method. \ The
importance of the influence of $Z\ $was observed in the mean energy $%
\overline{E}_{\alpha }\ $and angle $\overline{\theta }_{\alpha }.$ For a
given target $\alpha ,$ as $Z$ increases $\overline{E}_{\alpha }$ increases
but $\overline{\theta }_{\alpha }$ decreases \ At high impact energy, say
larger than 1 MeV/amu these values tend to the ones of the Born
approximation which embodies the simple Z$^{2}$ law. Sixteen molecules were
investigated using the simple stoichiometric model. Results for the six DNA
basis were presented and compared with the sparse available experiments \ We
explore the rule of Toburen which scales all the molecular ionization cross
section when divided by the number of weakly bound valence electrons $\nu
_{\alpha }$ given by Eq.(\ref{27})$.$ We have found the rule scales much
better when normalizing our theoretical ionization cross sections to the
number\ $\upsilon _{\alpha }^{\prime }$ given by Eq.(\ref{35}). And finally
we attempt to improve the stoichiometric model by the use of the Mulliken
charge to redefine a new stoichiometric model containing continuum rather
than integer proportions. NO substantial correction was found indicating
that the SSM works quite well.

Our aims is this article is to provide the tools to estimate any inelastic
parameter parameter -such as the emission angle, the emitted mean energy and
cross section- by the impact of any multicharged \ on any molecule
containing H, C, N, O, P and S, with the help of the stoichimetrical model.
Our goal was quite \textbf{pretencioso}, considering the simplicity of our
proposal. However we think our results could be used to estimate the
ionization magnitude with an acceptable level of uncertailties

@@

\section{References}

@@

>>>>

The first attempt to develop a comprehensive but straightforward 
phenomenological model for electron ejection from large molecules was 
proposed by Toburen {\it et al.}~\cite{toburen1975,toburen1976}. 
The authors found it convenient to scale the experimental ionization 
cross section in terms of the number of outer or weakly bound valence 
electrons (total number of electrons minus the inner-shell ones). 
We introduce the cross section per weakly bound electron or ionization 
cross section per valence electron, $\sigma_{e}$, as
\begin{equation}
\sigma_{e}=\frac{\sigma_{M}}{N_{M}}=\frac{\sum\limits_{\alpha}
n_{\alpha}\sigma_{\alpha}}{\sum\limits_{\alpha}n_{\alpha}\nu_{\alpha}}
=\sigma_{e}(v)\,, 
\label{27} 
\end{equation}
where
\begin{equation}
\nu_{\alpha}=\left\{ 
\begin{array}{ll}
4, & \text{for C,} \\ 
5, & \text{for N and P,} \\ 
6, & \text{for O and S,}
\end{array}\right.
\label{eq:nelec} 
\end{equation}
and $\nu_{\text{H}}=1$. The Toburen rule can be stated by saying that 
$\sigma_{e}$ is a \textit{universal} parameter independent on the 
molecule, which depends solely on the impact velocity. 
In Ref.~\cite{toburen1976}, it was found experimentally that for proton 
impact on some simple molecules, $\sigma_{e}$ results
\begin{equation}
\begin{array}{cccc}
E\text{(MeV)}               & 0.25 & 1.00 & 2.00 \\ 
\sigma_{e} (10^{-16}cm^{2}) & 0.39 & 0.17 & 0.11
\end{array}
\label{30}
\end{equation}
with an estimated error of about 20\%. A similar ratio was found in 
Ref.~\cite{itoh2013} for proton impact on uracil and adenine. 
Figure 4a shows $\sigma_{e}$ calculated with the CDW as a function of 
the impact velocity for different projectile charges computed with the 
SSM for the sixteen molecular targets displayed in 
Table~\ref{tab:families}. The universality with $Z$ is the one provided 
by Born approximation, i.e., $\sigma _{e}(Z)=Z^{2}\sigma_{e}(Z=1)$, 
and it holds for large impact velocities, as shown Figure 4a.
Of course, for lower impact velocities, the CDW breaks the behavior of 
the $Z^{2}$ rule. Although the Toburen rule holds for high energies, 
its performance is still not satisfactory: the universal band is quite 
broad. The departure of our theoretical 
results from the Toburen rule can be easily explained by the fact that 
our atomic cross sections $\sigma_{\alpha}$ behave differently for 
each atom in terms of the projectile charges. By inspecting Figure 1, 
one can easily see that the rule $\sigma_{\alpha}/\nu_{\alpha}\sim
\sigma_{e}$ is not well satisfied by the CDW. A much better general 
atomic rule is given by
\begin{equation}
\sigma_{e}^{\text{CDW}}=\frac{\sum\limits_{\alpha}
n_{\alpha}\sigma_{\alpha}^{\text{CDW}}}{\sum\limits_{\alpha}n_{\alpha}
\nu_{\alpha}^{\text{CDW}}},
\label{32} 
\end{equation}
where 
\begin{equation}
\nu_{\alpha }^{\text{CDW}} \sim\left\{ 
\begin{array}{ll}
4,\ \ \  & \text{for C, N, and O,} \\ 
4.5,\ \  & \text{for P and S}
\end{array}
\right. 
\label{eq:scalingCDW}
\end{equation}
and, obviously, $\nu_{\text{H}}^{\text{CDW}}=1$. The cross sections 
$\sigma_{e}^{\text{CDW}}$ are plotted in Figure~4b. A much better sharp 
band is observed, especially at high energy, where the theory is 
expected to work. It will be interesting for experiments to cross-check 
this theoretical prediction.

@@

In Figure 8, we display the relative difference of ionization cross 
sections 
\begin{equation}
e=\frac{\sigma^{\prime}-\sigma}{\sigma}\times 100
\label{120}
\end{equation}
for the four DNA bases. The difference obtained is of just very few 
percents, which indicates that the SSM is a quite robust model to treat 
the type of molecule we are dealing with, and of course within the 
range error that one can expect from the SSM.

Our molecular binding energies of the outer electrons for adenine, 
cytosine, guanine and thymine are shown in Figure~7. On the left side 
of the figure, we show the atomic Hartree--Fock energies of the 
constituent elements, which can give an idea of how the molecular levels 
distribute. A dashed line around -0.8 a.u. is drawn to separate in two 
the band levels of the outer electrons. The electrons in the molecular 
targets corresponding to the $2s$ of N and C in the atomic case are 
placed in a secondary gap. The lower levels of the secondary gap 
correspond to the oxygen. The number of electrons in the valence 
shell for all O would seem to be equal to 4, as previously setted by the 
scaling rule Eq.~(\ref{eq:scalingCDW}). The N case is not as straightfoward;
the electrons that would correspond to the $2s$ are located just below
the line. The $\nu_{\alpha }^{\text{CDW}}$ value given for N would 
suggest that a significant amount of electron density from the secondary gap 
is shared with the valence band.

\begin{figure}[H]
\centering
\includegraphics[width=0.95\textwidth]{energylevels/DNA.eps}
\end{figure}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

We now return to the calculation of the mean electron energy of H$^{+}$ 
and He$^{2+}$ in water, $\overline{E}_{\text{H}_2\text{O}}$. If we 
consider that the residual charge of each hydrogen in water is 
$q_{\text{H}}=+0.35$~\cite{rappe1991}, which implies that on oxygen we have a 
residual charge of $q_{\text{O}}=-0.7$, we could re-write the formula of water 
using Eq.~(\ref{120}) as H$_{1.3}$O$_{1.17}$ by considering 
$n_{\text{O}}^{\text{CDW}}\simeq 4$ and 
$n_{\text{H}}^{\text{CDW}}\simeq 1$. Recalculating mean electron energy 
by impact of H$^{+}$ and He$^{2+}$ on water at 1 MeV/amu, we obtain
$\overline{E}_{\text{H}_2\text{O}}=46.2$ and 47.6 eV. Our results 
are 10\% apart from the experimental values: 51.5 and 
52.2~eV~\cite{pimblott2007}, respectively, which is not bad considering 
the simplicity of our model.

Another parameter that can computed within our model is the fraction of
energy carried by the electron $f$, which is defined as
\begin{equation}
f=\frac{\sum\limits_{\alpha}n_{\alpha}\sigma_{\alpha}
\overline{E_{\alpha}}}{\sum\limits_{\alpha}n_{\alpha}(\sigma_{\alpha}
\overline{E_{\alpha}}+\sum\limits_{nl}\sigma_{\alpha nl}E_{\alpha nl})}\,.
\label{90}
\end{equation}
Using our shell-to-shell ionization cross sections 
$\sigma_{\alpha nl}$~\cite{miraglia2019}, and atomic binding energies of 
the atom $\alpha$, $E_{\alpha nl}$~\cite{clementi}, we obtain 
$f=0.70$ and $0.71$ on water at 1 MeV/amu for H$^{+}$ and He$^{2+}$, 
which differ from the respective experimental values $f=0.81$ and $0.80$.

The difference between our calculations and the experiments could be 
attributed to the binding energy values of the molecular water~12.65 eV 
and its atomic components, H (13.6 eV) and O (17.19 eV). This theoretical
discrepancy goes beyond our independent atom approximation and so it 
cannot be accounted for by our model. Once again, we could estimate this 
contribution by resorting to the stoichiometric model. We can make an 
asses for a maximum difference per molecule of water as
\begin{equation}
\delta E=(13.6-12.65)\times1.17+(17.19-12.65)\times1.3=7\,\text{eV}  
\label{80}
\end{equation}
Most of this energy should be transfer to the kinetic energy of the 
emitted electron. Thus, by adding 7 eV to the previous values, we obtain
$\overline{E}_{\text{H}_2\text{O}}=53.2$ and 54.6 eV for H$^{+}$ and 
He$^{2+}$ on water at 1 MeV/amu, respectively, which are much closer to
the experimental values. If we add $\delta E$ to the numerator of 
Eq.~(\ref{90}), we obtain $f=0.81$ and $0.80$, in perfect agreement with 
the experiments (see Table 1 of Ref.~\cite{pimblott2007}).

\section{Conclusions}

We calculated ionization cross sections by impact of antiprotons, 
H$^{+}$, He$^{2+}$, Be$^{4+}$, C$^{6+}$, and O$^{8+}$ for sixteen 
molecules containing H, C, N, O, P and S with the CDW method. The 
importance of the influence of $Z$ was observed in the mean energy 
$\overline{E}_{\alpha}$ and angle $\overline{\theta}_{\alpha}$. 
For a given target $\alpha$, as $Z$ increases, $\overline{E}_{\alpha}$ 
increases, but $\overline{\theta}_{\alpha}$ decreases. At high impact 
energy, say larger than 1 MeV/amu, these values converge to the Born 
approximation, which embodies the simple $Z^{2}$ law. The sixteen 
molecules selected were investigated using the simple stoichiometric 
model. Results for six DNA bases were presented and compared with the 
sparse available experiments. We explore the rule of Toburen which scales 
all the molecular ionization cross section when divided by the number of 
weakly bound valence electrons $\nu_{\alpha}$ given by Eq.~(\ref{27}). 
We found the rule scales much better when normalizing our theoretical 
ionization cross sections to the number $\nu_{\alpha}^{\text{CDW}}$ 
given by Eq.~(\ref{eq:scalingCDW}). Finally, we attempt to improve the 
stoichiometric model by using the Mulliken charge to define a new model 
containing fractional rather than integer proportions. No substantial 
correction was found, which indicates that the SSM works quite well.
By inspecting the molecular binding energy from quantum mechanical 
structure calculations, we were able to understand the values defined
for $\nu_{\alpha}^{\text{CDW}}$.

The main objective of this article is to provide the tools to estimate 
any inelastic parameter parameter --such as emission angle, emitted mean 
energy and cross section-- by the impact of any multicharged on any 
molecule containing H, C, N, O, P and S, with the help of the 
stoichimetrical model. Our goal was quite ambitious, considering the 
simplicity of our proposal. However, we think our results could be used 
to estimate the ionization magnitude with an acceptable level of 
uncertainty.

@@

@@

\bibitem{clementi}
E. Clementi, C. Roetti,
At. Data Nucl. Data Tables 14, 177--478 (1974).
